Normalization by Evaluation for Sized Dependent Types

### 2017-02-25 nbe/icfp17.tex
[ ICFP 17 ] case and fix

@@

\fbox{$w @ e \evalsto w'$}
\begin{gather*}
  \ru{t[u] \evalsto w
    }{(\lambda t) @ u \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) @ \alpha \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) @ \ann \alpha \evalsto w}
\\[2ex]
  \ru{t_z \evalsto w
    }{(\zero \alpha) @ \casett \evalsto w}
\qquad
  \ru{t_s\, t \evalsto w
    }{(\suc \alpha t) @ \casett \evalsto w}
\\[2ex]
  \ru{t\,\ann \alpha\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
     }{c @ \fixtt \evalsto w
     }{c\in \{ \zero \alpha, \suc \alpha u \}}
\end{gather*}
The rules added for NbE deal with elimination of delayed reflection:
\begin{gather*}
  \ru{T' \evalsto \piT U T
    }{(\up {T'} n) @ u \evalsto \up{T[u]} (n \, \down U u) }
\qquad
  \ru{T' \evalsto \piT \Size T
    }{(\up {T'} n) @ \alpha \evalsto \up{T[\alpha]} (n \, \alpha)}
\qquad
  \ru{T' \evalsto \forallT T
    }{(\up {T'} n) @ \ann \alpha \evalsto \up{T[\alpha]} (n \ann \alpha)}
%   \ru{T \evalsto \piT A B \qquad
%       \up{B[u]} (n \, \down A u) \evalsto w
%     }{(\up T n) @ u \evalsto w}
% \qquad
%   \ru{T \evalsto \piT \Size B \qquad
%       \up{B[u]} (n \, \alpha) \evalsto w
%     }{(\up T n) @ \alpha \evalsto w}
% \qquad
%   \ru{T \evalsto \forallT B \qquad
%       \up{B[\alpha]} (n \ann \alpha) \evalsto w
%     }{(\up T n) @ \ann \alpha \evalsto w}
\\[2ex]
  \ru{T' \evalsto \Nat \beta \qquad
    }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
      \up{T\,(\up{\Nat \infty} n)}
         n\,\case[\ell]
             T %% {(\down {\Set[\infty]} T)}
             {(\down {T\,{\zero\infty}} t_z)}
             {(\down {\funS x {\Nat \infty} T\,(\suc \infty x)} t_s)}
    }
\\[2ex]
  \ru{T' \evalsto \Nat \beta \qquad
    }{(\up {T'} n) @ \fix[\ell] T t \evalsto
      n\,\fix[\ell] {(\down {\FixK} T)} {(\down {\FixT\, T} t)}
    }
\end{gather*}

@@

\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{%t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k {\lev j\,\vec e} {\ind{k \monus (1+j)}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k d v
    }{\RElim k d v}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k \alpha b}
\qquad
  \ru{
    }{\RElim k {\ann \alpha} {\ann \infty}}
  % \ru{\RSize k \alpha b
  %   }{\RElim k {\ann \alpha} {\ann b}}
\\[2ex]
  \ru{\RTy k T V \qquad
      \R k {d_z} {v_z} \qquad
      \R k {d_s} {v_s}
    }{\RElim k {(\case[\ell] T {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\R k D V \qquad
      \R k d v
    }{\RElim k {(\fix[\ell] D d)} {\fix[\ell] V v}}
\end{gather*}
\fbox{$\RTy k T V$}
\begin{gather*}
  \ru{T \evalsto \Set[\ell]
    }{\RTy k T {\Set[\ell]}}
\qquad
  \ru{T \evalsto \Nat \alpha \qquad
      \RSize k \alpha b
    }{\RTy k T {\Nat b}}
\qquad
  \ru{% T \evalsto \up{U}n \qquad \RNe k n m
      \RUp k T m
    }{\RTy k T m}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      \RTy k A {V_a} \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\piT {V_a} V_b}}
\qquad
  \ru{T \evalsto \epiT \Size B \qquad
      \RTy {k+1} B {V}
    }{\RTy k T {\epiT \Size V}}
\end{gather*}

@@

@@

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is PER of values of type $A$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{% T \evalsto \up{\Set[\ell']} n \qquad
      % T' \evalsto \up{\Set[\ell'']} n' \qquad
      % n = n' \in \NE
      T = T' \in \NEE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NEE
\\[3ex]
  \ru{T \evalsto \Nat\,\alpha \qquad T' \evalsto \Nat\,\alpha
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(\alpha)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; \alpha \mapsto \EL[\ell](B[\alpha]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(\alpha \mapsto \EL[\ell](B[\alpha]))
\end{array}
\]
All relations involved here are closed under weak head expansion.
\begin{lemma}[Well-definedness]
  Let $\DD :: T_1 = T_2 \in \SET[\ell]$.
  \begin{enumerate}
  \item Symmetry: $T_2 = T_1 \in \SET[\ell]$.
  \item Transitivity: If $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
  \item Extension: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Simultaneously by induction on $\DD$.
\end{proof}

@@

>>>>

\makeatletter
\newenvironment{proof*}[1][\proofname]{\par
  \normalfont \topsep6\p@\@plus6\p@\relax
  \trivlist
  \item[\@proofindent\hskip\labelsep
        {\@proofnamefont #1\@addpunct{.}}]\ignorespaces
}{%
  \endtrivlist\@endpefalse
}
\makeatother

@@

\fbox{$w @ e \evalsto w'$}
\begin{gather*}
  \ru{t[u] \evalsto w
    }{(\lambda t) @ u \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) @ \alpha \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) @ \ann \alpha \evalsto w}
\\[2ex]
  \ru{t_z \evalsto w
    }{(\zero \alpha) @ \casett \evalsto w}
\qquad
  \ru{t_s\, t \evalsto w
    }{(\suc \alpha t) @ \casett \evalsto w}
\\[2ex]
  \ru{t\,\ann \alpha\,(\lambda x.\, x\, \fixtt)\,c \evalsto w
     }{c @ \fixtt \evalsto w
     }{c\in \{ \zero \alpha, \suc \alpha u \}}
\end{gather*}
The rules added for NbE deal with elimination of delayed reflection:
\begin{gather*}
  \ru{T' \evalsto \piT U T
    }{(\up {T'} n) @ u \evalsto \up{T[u]} (n \, \down U u) }
\qquad
  \ru{T' \evalsto \piT \Size T
    }{(\up {T'} n) @ \alpha \evalsto \up{T[\alpha]} (n \, \alpha)}
\qquad
  \ru{T' \evalsto \forallT T
    }{(\up {T'} n) @ \ann \alpha \evalsto \up{T[\alpha]} (n \ann \alpha)}
%   \ru{T \evalsto \piT A B \qquad
%       \up{B[u]} (n \, \down A u) \evalsto w
%     }{(\up T n) @ u \evalsto w}
% \qquad
%   \ru{T \evalsto \piT \Size B \qquad
%       \up{B[u]} (n \, \alpha) \evalsto w
%     }{(\up T n) @ \alpha \evalsto w}
% \qquad
%   \ru{T \evalsto \forallT B \qquad
%       \up{B[\alpha]} (n \ann \alpha) \evalsto w
%     }{(\up T n) @ \ann \alpha \evalsto w}
\\[2ex]
  \ru{%T' \evalsto \Nat \beta \qquad
    }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
      \up{T\,(\up{T'} n)}\,
      % \up{T\,(\up{\Nat \infty} n)}
         n\,\case[\ell]
             {(\down {\Nat \infty \to \Set[\ell]} T)}
             {(\down {T\,{\zero\infty}} t_z)}
             {(\down {\funS x {\Nat \infty} T\,(\suc \infty x)} t_s)}
    }
\\[2ex]
  \ru{%T' \evalsto \Nat \beta \qquad
    }{(\up {T'} n) @ \fix[\ell] T t \evalsto
      n\,\fix[\ell] {(\down {\FixK} T)} {(\down {\FixT\, T} t)}
    }
\end{gather*}

@@

\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{%t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k {\lev j\,\vec e} {\ind{k \monus (1+j)}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k d v
    }{\RElim k d v}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k \alpha b}
\qquad
  \ru{
    }{\RElim k {\ann \alpha} {\ann \infty}}
  % \ru{\RSize k \alpha b
  %   }{\RElim k {\ann \alpha} {\ann b}}
\\[2ex]
  \ru{\R k D V \qquad
      \R k {d_z} {v_z} \qquad
      \R k {d_s} {v_s}
    }{\RElim k {(\case[\ell] D {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\R k D V \qquad
      \R k d v
    }{\RElim k {(\fix[\ell] D d)} {\fix[\ell] V v}}
\end{gather*}
\fbox{$\RTy k T V$}
\begin{gather*}
  \ru{T \evalsto \Set[\ell]
    }{\RTy k T {\Set[\ell]}}
\qquad
  \ru{T \evalsto \Nat \alpha \qquad
      \RSize k \alpha b
    }{\RTy k T {\Nat b}}
\qquad
  \ru{% T \evalsto \up{U}n \qquad \RNe k n m
      \RUp k T m
    }{\RTy k T m}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      \RTy k A {V_a} \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\piT {V_a} V_b}}
\qquad
  \ru{T \evalsto \epiT \Size B \qquad
      \RTy {k+1} B {V}
    }{\RTy k T {\epiT \Size V}}
\end{gather*}

@@

Now we define some PERs and PER constructors on values.  All these PERs $\A$ are closed under weak head equality, meaning if $a = b \in \A$ and $a'$ has the same weak head normal form as $a$, then $a' = b in \A$.  (By symmetry, $\A$ is also closed under weak head equality on the second argument.)

@@

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is PER of values of type $A$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{% T \evalsto \up{\Set[\ell']} n \qquad
      % T' \evalsto \up{\Set[\ell'']} n' \qquad
      % n = n' \in \NE
      T = T' \in \NEE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NEE
\\[3ex]
  \ru{T \evalsto \Nat\,\alpha \qquad T' \evalsto \Nat\,\alpha
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(\alpha)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; \alpha \mapsto \EL[\ell](B[\alpha]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(\alpha \mapsto \EL[\ell](B[\alpha]))
\end{array}
\]
All relations involved here are closed under weak head equality.
\begin{lemma}[Well-definedness]
  Let $\DD :: T_1 = T_2 \in \SET[\ell]$.
  \begin{enumerate}
  \item Symmetry: $T_2 = T_1 \in \SET[\ell]$.
  \item Transitivity: If $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
  \item Extension: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Simultaneously by induction on $\DD$.
\end{proof}

@@

\subsection{Computation}
\label{sec:comp}

The eliminations for natural numbers are acurrately modeled.
\begin{lemma}[Case]
  \label{lem:case}
If $a = a' \in \Nat (\alpha+1)$
and $B = B' \in \Nat (\alpha+1) \to \Set[\ell]$
and $f_z = f_z' \in B\,(\zero \beta)$
and $f_s = f_s' \in \funT x {\Nat \alpha} B\,(\suc \gamma x)$
then $a\,\case[\ell] B {f_z}{f_s} = a'\,\case[\ell] {B'} {f_z'}{f_s'} \in B\,a$.
\end{lemma}
\begin{proof}
By induction on $a = a' \in \Nat(\alpha+1)$.
\begin{caselist}

\nextcase $a \evalsto \zero \beta$ and $a' \evalsto \zero{\beta'}$.
Since our PERs are closed under weak head equality,
and, for instance, $a\, \case[\ell] B {f_z}{f_s}$ has the same weak head normal form as $f_z$,
it suffices to show $f_z = f_z' \in B\,(\zero \beta)$, which is one of our assumptions.

\nextcase $a \evalsto \suc \beta b$ and $a' \evalsto \suc {\beta'}{b'}$ with $b = b' \in \Nat \alpha$.
Again, it suffices to show
$f_s\,b = f_s'\,b' \in B\,(\suc \beta b)$, which is an instance of our last assumption.

\nextcase $a \evalsto \up T n$ and $a' \evalsto \up {T'} n'$ with $n = n' \in \NE$.
Let $D = \down{\Nat \infty \to \Set[\ell]} B$
and $d_z = \down{B\,(\zero \infty)} f_z$
and $d_s = \down{\funS x {\Nat\,\infty} {B\,(\suc \infty x)}} f_s$
and $e = \case[\ell] D {d_z}{d_s}$.
% $e = \case[\ell]
%   {(\down{\Nat \infty \to \Set[\ell]} B)}
%   {(\down{B\,(\zero \infty)} f_z)}
%   {(\down{\funS x {\Nat\,\infty} {B\,(\suc \infty x)} f_s)}$
Let $D',d_z',d_s',e'$ be defined analogously from $B',f_z',f_s'$.  It suffices to show $e = e' \in \ELIM$, since then we have $n\,e = n'\,e' \in \NE$ by the closure properties of $\NE$ (Lemma~\ref{lem:closne}),
and $\up{B[a]} (n\,e) = \up{B'[a']} (n'\,e') \in B[a]$ by reflection (Theorem~\ref{thm:rere}).
The remaining goal $\case[\ell] D {d_z}{d_s} = \case[\ell] {D'} {d_z'}{d_s'} \in \ELIM$ follows by the closure properties for eliminations (Lemma~\ref{lem:closelim}), since $D = D' \in \NF$ and $d_z = d_z' \in \NF$ and $d_s = d_s' \in \NF$ all hold by reification (Theorem~\ref{thm:rere}).
\end{caselist}
\end{proof}

\begin{lemma}[$\NAT$ is cocontinuous]
  \label{lem:limit}
  $\NAT(\infty) = \bigcup_{\alpha < \infty} \NAT(\alpha)$.
\end{lemma}
\begin{proof}
  By induction on $a = a' \in \NAT(\infty)$, we can easily show $a = a' \in \NAT(\alpha)$ for some $\alpha < \infty$.
  For instance, $\alpha$ could be number of uses of the successor rule plus one.
\end{proof}

%A type constructor $B \in \FixK\,\ell$ is \emph{admissible for recursion}, \fbox{$B \in \ADM$}, if
%$\bigcap_{
Let us write \fbox{$B = B' \in \Adm\ell$} iff $B = B' \in \FixK\,\ell$ and
for all $\beta \in \SIZE$ and $a \in \Nat \beta$ we have
$B\,\beta\, a \leq B\,\infty\,a \in \TYPE$ and
$B'\,\beta\, a \leq B'\,\infty\,a \in \TYPE$.
\begin{lemma}[Fix]
Let $g = a \, \fix[\ell] B f$ and $g' = a'\,\fix[\ell]{B'}{f'}$.
If $a = a' \in \Nat \alpha$
and $B = B' \in \Adm\ell$
and $f = f' \in \FixT\,B$
then $g = g' \in B\,\alpha\,a$.
\end{lemma}
\begin{proof*}
By wellfounded induction on $\alpha$.
\begin{caselist}

\nextcase $\alpha < \infty$ and $a \evalsto \up{T}n$ and $a' \evalsto \up{T'}{n'}$ and $n = n' \in \NE$.
In this case $g$ and $g'$ evaluate to neutral applications of $\tfix$.  The proof proceeds analogously to Lemma~\ref{lem:case}.

\nextcase $\alpha < \infty$ and $a \evalsto c$ and $a' \evalsto c'$.  Then $\alpha = \beta + 1$ with $\beta<\infty$.
The weak head normal form of $g$ equals the weak head normal form of
$h := f\,\ann \gamma\,(\lambda x.\,x\,\fix[\ell] B f)\,c$ where $\gamma$ is the size annotation of $c$.
It suffices to show $h = h' \in B\,\alpha\,a$ for $h'$ defined analogously from $B',f',c'$.
This follows from the assumption on $f, f'$ if we manage to show
$(\lambda x.\,x\,\fix[\ell] B f) = (\lambda x.\,x\,\fix[\ell] {B'} {f'}) \in
\funT x {\Nat\beta} {B\,\beta\,x}$.
To this end, assume $b = b' \in \Nat\beta$ and show
$b\,\fix[\ell] B f = b'\,\fix[\ell] {B'} {f'} \in B\,\beta\,b$.
However, this is an instance of the induction hypothesis thanks to $\beta < \alpha$.

\nextcase $\alpha = \infty$.
Note that $a = a' \in \Nat \alpha$ for some $\alpha < \infty$ by Lemma~\ref{lem:limit}.
By induction hypothesis, $g = g' \in B\,\alpha\,a$.  Since
$B\,\alpha\,a \leq B\,\infty\,a$ by assumption, the goal
$g = g' \in B\,\infty\,a$ follows by subsumption.
\end{caselist}
\end{proof*}
